### 2020-06-18 report/sections/evalconcl/eval/game.tex
Report: final proofread

new Svr(
	wss,						// WebSocket server
	handleP1Move,		// Game logic
	cancellation		// Error handler
);
\end{lstlisting}
\captionof{lstlisting}{Implementing \tprotocol{Noughts and Crosses} Game Server using
\codegen}
\label{lst:gamesvr}
\end{figure}

@@

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
// Inside some render() function...
{board.map((row, x) =>
	<tr>
	{row.map((cell, y) => {
		if (cell === Cells.VACANT) {
			const sendPoint = (event: React.MouseEvent) => {
				// Don't refresh the screen
				event.preventDefault();
				return { x, y };
			});
			const SelectPoint = this.pros.Pos('onClick', sendPoint);
			return <SelectPoint><td>{cell}</td></SelectPoint>
		} else {
			// Render nought or cross,
			// but clicking on this cell will *not* send anything
			return <td>{cell}</td>
		}		
	})}
	</tr>
)}
\end{lstlisting}
\captionof{lstlisting}{Safely Binding Send Actions to 
\tprotocol{Noughts and Crosses} Game Board}
\label{lst:gamesendfactory}
\end{figure}

>>>>

new Svr(
	wss,						// WebSocket server
	handleP1Move,		// Game logic
	cancellation		// Error handler
);
\end{lstlisting}
\captionof{lstlisting}{Implementing \tprotocol{Noughts and Crosses} Game Server}
\label{lst:gamesvr}
\end{figure}

@@

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
// Inside some render() function...
{board.map((row, x) =>
	<tr>
	{row.map((cell, y) => {
		if (cell === Cells.VACANT) {
			const sendPoint = (event: React.MouseEvent) => {
				return { x, y };
			});
			const SelectPoint = this.pros.Pos('onClick', sendPoint);
			return <SelectPoint><td>{cell}</td></SelectPoint>
		} else {
			// Render nought or cross,
			// but clicking on this cell will *not* send anything
			return <td>{cell}</td>
		}		
	})}
	</tr>
)}
\end{lstlisting}
\captionof{lstlisting}{Safely Binding Send Actions to 
\tprotocol{Noughts and Crosses} Game Board}
\label{lst:gamesendfactory}
\end{figure}

### 2020-06-18 report/sections/general/p2p/approach.tex
Report: final proofread

We acknowledge that peer-to-peer browser interactions
could be achieved via different transport abstractions, 
such as the WebRTC framework \cite{WebRTC}. 
However, WebRTC has its own proprietary ``protocol'' 
for establishing connections and exchanging network information,
which isn't formalised in session type theory. This setup phase
generally requires a \textit{signalling server} \cite{WebRTCSignaling},
which brings our problem full circle, as it can be implemented
over WebSocket transport. Extending API generation to also include
the boilerplate for initialising WebRTC connections could be an option,
but would introduce complications in the session runtime and
error handling (e.g. how to handle cancellations during the connection
phase?).
We believe that adopting WebRTC deviates from the focus of project in 
implementing type-safe webservices.

>>>>

We acknowledge that peer-to-peer browser interactions
could be achieved via different transport abstractions, 
such as the WebRTC framework \cite{WebRTC}. 
However, WebRTC has its own ``protocol'' 
for establishing connections and exchanging network information,
which isn't formalised in session type theory. This setup phase
generally requires a \textit{signalling server} \cite{WebRTCSignaling},
which brings our problem full circle, as it can be implemented
over WebSocket transport. Extending API generation to also include
the boilerplate for initialising WebRTC connections could be an option,
but would introduce complications in the session runtime and
error handling (e.g. how to handle cancellations during the connection
phase?).
We believe that adopting WebRTC deviates from the focus of project in 
implementing type-safe web services.

### 2020-06-18 report/sections/general/p2p/challenges.tex
Report: final proofread

\begin{example}[Naive definition of routing in global types]
Consider the global type
\[
A \to B: \tmsg{M1}. S \to B: \tmsg{M2}. \texttt{end}
\]
By existing LTS semantics over
global types \cite{characterisation}, 
then $\aout{S}{B}{\tmsg{M2}}$
is a prefix of a valid execution trace.
If we define our routing construct simply by replacing every single
$A \to B: \tmsg{M}$ interaction with $A \to S: \tmsg{M}.
S \to B: \tmsg{M}$, we transform the global type into
\[
A \to S: \tmsg{M1}. S \to B: \tmsg{M1}. S \to B: \tmsg{M2}. \texttt{end}
\]
our naive definition of routing has over-serialised the protocol
since $\aout{S}{B}{\tmsg{M2}}$ can no longer occur first because \trole{S}
is forced to send \tmsg{M1} first, which it cannot do so until receiving
from \trole{A}.
This implies that we need to handle routing interactions differently
from ``normal'' send/receive actions.
\end{example}

There is existing work \cite{LinearDecomp, BinaryDecomp, BinaryDuality}
that resembles our routing proposal through
\textit{decomposing} multiparty sessions into a collection
of binary sessions,
but these approaches
specify an endpoint to exclusively carry out
routing responsibilities and act as a centralised orchestrator 
for the multiparty session.
Our proposal recognises that, in the context of web applications,
there is a need for server-side endpoints to implement business logic
so enforcing the server to be an exclusive orchestrator makes our work
incompatible with webservice protocols; 
we also argue that the server can still participate in the 
multiparty session and concurrently perform its routing duties
in a way that is transparent to non-server endpoints, and crucially,
does not overserialise the permitted interactions. 
We introduce our extension as \textit{Routed Session Types},
and proceed to prove these claims in \cref{chap:theory}.

Likewise, the implementation of routed session types 
in \fancyname{SessionTS} needs to accurately reflect the theory
and be as transparent to the developer as possible to minimise
the learning curve overhead demanded 
by complicating the generated APIs -- we document our extensions
to \fancyname{SessionTS} in \cref{chap:impl}.

>>>>

\begin{example}[Naive definition of routing in global types]
Consider the global type
\[
A \to B: \tmsg{M1}. S \to B: \tmsg{M2}. \texttt{end}
\]
By existing LTS semantics over
global types introduced in 
Definition 3.2 of \cite{characterisation}, 
then $\aout{S}{B}{\tmsg{M2}}$
is a prefix of a valid execution trace.
If we define our routing construct simply by replacing every single
$A \to B: \tmsg{M}$ interaction with $A \to S: \tmsg{M}.
S \to B: \tmsg{M}$, we transform the global type into
\[
A \to S: \tmsg{M1}. S \to B: \tmsg{M1}. S \to B: \tmsg{M2}. \texttt{end}
\]
our naive definition of routing has over-serialised the protocol
since $\aout{S}{B}{\tmsg{M2}}$ can no longer occur first because \trole{S}
is forced to send \tmsg{M1} first, which it cannot do so until receiving
from \trole{A}.
This implies that we need to handle routing interactions differently
from ``normal'' send/receive actions.
\end{example}

There is existing work \cite{LinearDecomp, BinaryDecomp, BinaryDuality}
that resembles our routing proposal through
\textit{decomposing} multiparty sessions into a collection
of binary sessions,
but these approaches
specify an endpoint to exclusively carry out
routing responsibilities and act as a centralised orchestrator 
for the multiparty session.
Our proposal recognises that, in the context of web applications,
\textit{there is a need for server-side endpoints to 
implement business logic},
so enforcing the server to be an exclusive orchestrator makes our work
incompatible with web service protocols; 
we also argue that the server can still participate in the 
multiparty session and concurrently perform its routing duties
in a way that is transparent to non-server endpoints, and crucially,
does not overserialise the permitted interactions. 
We introduce our extension as \newtheory,
and proceed to prove these claims in \cref{chap:theory}.

Likewise, the implementation of routed session types 
in \codegen needs to accurately reflect the theory.
It also needs to be 
as transparent to the developer as possible to minimise
the learning curve overhead demanded 
by complicating the generated APIs.
We document our extensions
to \codegen in \cref{chap:impl}.

### 2020-06-18 report/sections/general/theory.tex
Report: final proofread

\newtheory provide the theoretical basis
for allowing developers to implement
protocols with browser-to-browser interactions
(such as the \tprotocol{Two Buyer} protocol
introduced in \cref{section:twobuyer})
over a server-centric network topology
which uses WebSocket transport.
We introduce the extended syntax for types in 
\cref{section:syntax}, and give semantics of our
syntax extensions in \cref{section:lts}.
We define a router-parameterised encoding
for MPST into \newtheory in \cref{section:encoding}
and prove the preservation of well-formedness
and communication.
We proceed to extend \codegen to implement \newtheory
in \cref{chap:impl}.

@@

\section{Summary}
We have presented \newtheory, a variant of
the canonical MPST theory to express \textit{routed communication}.
We introduced extensions to syntax and semantics, and proved
that our extended semantics preserve deadlock-freedom
for well-formed protocols. We defined an encoding from the
canonical theory onto \newtheory, and proved the preservation
of well-formedness and communication.

>>>>

\newtheory provide the theoretical basis
for allowing developers to implement
protocols with browser-to-browser interactions
(such as the \tprotocol{Two Buyer} protocol
introduced in \cref{section:twobuyer})
over a server-centric network topology
which uses WebSocket transport.
We introduce the extended syntax for types in 
\cref{section:syntax}, and give semantics of our
syntax extensions in \cref{section:lts}.
Regarding the extended semantics on routed session types,
we prove the soundness and completeness of
the extended semantics in \cref{section:newtraceeq}.
In \cref{section:encoding},
we define a router-parameterised encoding
for canonical MPST theory into \newtheory
and prove the preservation of well-formedness
and communication.
We proceed to extend \codegen to implement \newtheory
in \cref{chap:impl}.

@@

\section{Summary}
We have presented \newtheory, a variant of
the canonical MPST theory to express \textit{routed communication}.
We introduced extensions to syntax and semantics, and proved
that our extended semantics are sound and complete,
and preserve deadlock-freedom
for well-formed protocols. 
We defined an encoding from the
canonical theory onto \newtheory, and proved the preservation
of well-formedness and communication.

### 2020-06-18 report/sections/general/theory/lts.tex
Report: final proofread

We define the labelled transition system (LTS) semantics
over global types (\cref{subsection:newltsglobal}) 
and local types (\cref{subsection:newltslocal})
for \newtheory,
building upon the work of Deni\'elou and Yoshida in
\cite{characterisation}.
We show the soundness and completeness of projection
with respect to the LTSs through proving the
\textit{trace equivalence} of
a global type and the collection of local types projected
from the global type (\cref{subsection:newtraceeq}).
We then use this result to
conclude that \newtheory provide the same
communication safety guarantees from canonical
MPST theory for well-formed global types, 
namely deadlock freedom (\cref{subsection:newdeadlockfreedom}).

@@

\begin{itemize}
\item \textbf{Routed send:}

@@

\item \textbf{Routed receive:}

@@

We add a similar construct
{$\gtransroute{p}{q}{s}{j}{l_i: G_i}{i \in I}$}
to represent that the message $l_j$ has
been sent from $\role{p}$ to the router $\role{s}$
but not yet routed to $\role{q}$.
We extend the projection operator to 
support this new construct.

@@

@@

We define the LTS semantics 
over global types,
denoted by $\treduce{G}{G'}{l}$,
in \cref{fig:newglobal}.
We highlight and explain the new rules.

@@

The LTS semantics over local types
need to be extended for routed communication.
Because the router holds a ``global perspective''
on the routed communication, we also need to
represent the intermediate state (at which the message is 
being routed by the router) within the grammar
of local types.
Similar to \cref{subsection:newltsglobal}, we add 
the construct 
$\routertrans{p}{q}{j}{l_i: T_i}{i \in I}$
to represent that, from the local perspective of
the router, the message $l_j$
has been received from $\role{p}$ but not yet routed
to $\role{q}$.

>>>>

We define the labelled transition system (LTS) semantics
over global types (\cref{subsection:newltsglobal}) 
and local types (\cref{subsection:newltslocal})
for \newtheory,
building upon the work of Deni\'elou and Yoshida
\cite{characterisation}.
We show the soundness and completeness of projection
with respect to the LTSs through proving the
\textit{trace equivalence} of
a global type and the collection of local types projected
from the global type (\cref{subsection:newtraceeq}).
We then use this result to
conclude that \newtheory provide the same
communication safety guarantees from canonical
MPST theory for well-formed global types, 
namely deadlock freedom (\cref{subsection:newdeadlockfreedom}).

@@

\begin{itemize}
\item \textbf{Routed Send:}

@@

\item \textbf{Routed Receive:}

@@

We add a similar construct
{$\gtransroute{p}{q}{s}{j}{l_i: G_i}{i \in I}$}
to represent that the message $l_j$ has
been sent from $\role{p}$ to the router $\role{s}$
\textit{but not yet routed to $\role{q}$}.
We extend the projection operator to 
support this new construct.

@@

Because the router $\role{s}$ holds
Because the router holds a ``global perspective''
on the routed communication, we also need to
represent the intermediate state (at which the message is 
being routed by the router) within the grammar
of \textit{local types}.
We extend the grammar of local types with the construct
$\routertrans{p}{q}{j}{l_i: T_i}{i \in I}$
to represent that, from the local perspective of
the router, the message $l_j$
has been received from $\role{p}$ \textit{but not yet routed
to $\role{q}$}.

@@

We define the LTS semantics 
over global types,
denoted by $\treduce{G}{G'}{l}$,
in \cref{fig:newglobal}.
We highlight and explain the new rules.\\

@@

### 2020-06-18 report/sections/general/theory/syntax.tex
Report: final proofread

Global types range over $G, G', G_i, \dots$
As motivated by \cref{section:routerchallenges},
we need to handle \textit{routed communication} separately
from normal send and receive actions.
We define the syntax of global types for \newtheory 
in \cref{fig:newsyntaxglobal}, and explain the 
new construct for expressing routed communication.

@@

\item \textbf{Routed communication:}

@@

\item \textbf{Routed slection:}

@@

\item \textbf{Routed branching:}

@@

\doublespacing
\[
\begin{array}{rcl}
\tmerge
{(\tselproxy{p}{q}{l_i: T_i}{i \in I})}
{(\tselproxy{p}{q}{l_i: T_i}{i \in I})}
	& = & \tselproxy{p}{q}{l_i: T_i}{i \in I} \\

\tmerge
{(\tbraproxy{p}{q}{l_i: T_i}{i \in I})}
{(\tbraproxy{p}{q}{l_j: T'_j}{j \in J})}
	& = & \tbraproxy{p}{q}{l_k: T''_k}{k \in I \cup J} \\
\text{where} & & T''_k = \begin{cases}
T_k & \text{if} ~ k \in I \setminus J \\
T'_k & \text{if} ~ k \in J \setminus I \\
\tmerge{T_k}{T'_k} & \text{if} ~ k \in I \cap J \\
\text{otherwise} & \text{undefined}
\end{cases} \\
\end{array}
\]
\singlespacing

@@

In \newtheory, we require that there is \textit{exactly
one} router, say $\role{s}$, in the global type, so
we need to express that a global type to be
well-formed \textit{with respect to the role $\role{s}$
acting as the router}.

@@

\begin{itemize}
\item 
\rulename{$\centroidop$-Comm}:
For direct communication, $\role{s}$ must
be a participant and the centroid of all continuations.

\item 
\rulename{$\centroidop$-RoutedComm}:
For routed communication, $\role{s}$ must
be the router and be the centroid of all continuations.

@@

We also assume that $G$ has no free type variables
and is guarded/contractive.

>>>>

Global types range over $G, G', G_i, \dots$
As motivated by \cref{section:routerchallenges},
we need to handle \textit{routed communication} separately
from direct send and receive actions.
We define the syntax of global types for \newtheory 
in \cref{fig:newsyntaxglobal}, and explain the 
new construct for expressing routed communication.

@@

\item \textbf{Routed Communication:}

@@

\item \textbf{Routed Selection:}

@@

\item \textbf{Routed Branching:}

@@

\doublespacing
\[
\begin{array}{rl}
\tmerge
{\tselproxy{p}{q}{l_i: T_i}{i \in I}}
{\tselproxy{p}{q}{l_i: T_i}{i \in I}}
	& = \tselproxy{p}{q}{l_i: T_i}{i \in I} \\

\tmerge
{\tbraproxy{p}{q}{l_i: T_i}{i \in I}}
{\tbraproxy{p}{q}{l_j: T'_j}{j \in J}}
	& = \tbraproxy{p}{q}{l_k: T''_k}{k \in I \cup J} \\
\text{where} & T''_k = \begin{cases}
T_k & \text{if} ~ k \in I \setminus J \\
T'_k & \text{if} ~ k \in J \setminus I \\
\tmerge{T_k}{T'_k} & \text{if } ~ k \in I \cap J \\
\end{cases} \\
\text{otherwise} & \text{undefined}
\end{array}
\]
\singlespacing

@@

In \newtheory, we need to express that a global type is
well-formed \textit{with respect to the role $\role{s}$
acting as the router}.

@@

\begin{itemize}
\item 
\rulename{$\centroidop$-Comm}:
For direct communication, $\role{s}$ must
be a participant and a centroid of all continuations.

\item 
\rulename{$\centroidop$-RoutedComm}:
For routed communication, $\role{s}$ must
be the router and be a centroid of all continuations.

@@

We also assume that the syntax of $G$ is \textit{contractive}, 
i.e. that type variables are guarded.

### 2020-06-18 report/sections/introduction/contributions.tex
Report: final proofread

We present an implementation of the MPST framework for developing
interactive web applications over WebSocket transport.
We also present a theory of routed multiparty session types,
which forms the basis for supporting arbitrary communication structures
over a server-centric WebSocket-based network topology.
Related work \citep{PureScript2019, MVU2020, LINKS, Exceptional} 
requires the 
developer to implement their application using tools that are 
unconventional for web development, 
and only support communication protocols that adhere to the
WebSocket network structure.
To the best of our knowledge, this is the first work on multiparty session
type-safe web development that provides the same communication safety 
guarantees using familiar industry tools in TypeScript and React,
and supports peer-to-peer browser interactions over WebSockets.

In \textbf{\cref{chap:background}}, 
we introduce the core session type theory
and provide an overview of related work on integrating session types into
application development. We also introduce TypeScript and highlight the 
defining features of the language.

@@

In \textbf{\cref{chap:ext}},
we extend \codegen to support advanced
practices commonly found in full-stack web development in a way that
preserves communication safety. 
We modify the API generation strategies
to support \textit{asynchronous implementations} 
and introduce how to 
gracefully handle errors and \textit{session cancellation} 
in the context of interactive web applications.

@@

In \textbf{\cref{chap:p2p}}, 
we introduce the \tprotocol{Two Buyer} multiparty
protocol to motivate our proposal of providing the server role with 
routing responsibilities in addition to its existing interactions
specified in the protocol.

@@

\paragraph{\cref{part:evalconcl}}
In \textbf{\cref{chap:eval}}, 
we evaluate our work through 
case studies of communication protocols 
implemented as webservices. Through examples of
multiparty sessions and routed multiparty sessions, 
we demonstrate how the
developer may implement the generated APIs and 
consider the compatibility of our work with respect to
modern web programming practices.
We perform benchmarks on a web application built using our
generated APIs against a baseline implementation to evaluate
the overhead of our work.

>>>>

We present an implementation of the MPST framework for developing
interactive web applications over WebSocket transport.
We also present a theory of routed multiparty session types,
which forms the basis for supporting arbitrary communication structures
over a server-centric WebSocket-based network topology.

Related work \citep{PureScript2019, MVU2020, LINKS, Exceptional} 
requires the 
developer to implement their application using tools that are 
unconventional for web development, 
and only support communication protocols that adhere to the
WebSocket network structure.
To the best of our knowledge, this is the first work on multiparty session
type-safe web development that provides the same communication safety 
guarantees using familiar industry tools in TypeScript and React,
and supports peer-to-peer browser interactions over WebSockets. \\

In \textbf{\cref{chap:background}}, 
we introduce the core session type theory
and provide an overview of related work on integrating session types into
application development, with a focus on the state-of-the-art
proposals for session-typed web development. 
We also introduce TypeScript and highlight the 
defining features of the language.

@@

In \textbf{\cref{chap:ext}},
we extend \codegen to support advanced
practices commonly found in full-stack web development,
in a way that
preserves communication safety. 
We modify the API generation strategies
to support \textit{asynchronous implementations} 
and introduce how to 
gracefully handle errors and \textit{session cancellation} 
in the context of interactive web applications.

@@

In \textbf{\cref{chap:p2p}}, 
we introduce the \tprotocol{Two Buyer} multiparty
protocol to motivate our proposal of providing the server role with 
routing responsibilities, in addition to its existing interactions
specified in the protocol.

@@

\paragraph{\cref{part:evalconcl}}
In \textbf{\cref{chap:eval}}, 
we evaluate our work through 
case studies of communication protocols 
implemented as webservices. Through examples of
multiparty sessions and routed multiparty sessions, 
we demonstrate how the
developer may implement the generated APIs and 
consider the compatibility of our work with respect to
modern web programming practices.
We perform benchmarks on a web application built using our
generated APIs against a baseline implementation to evaluate
the overhead of our implementation.

### 2020-06-18 report/sections/introduction/motivation.tex
Report: final proofread

Consider a \textit{flight booking service}
between a \trole{Traveller} and a \trole{FlightServer}.
The \trole{Traveller} queries the \trole{FlightServer}
about prices to a particular destination.
If there is still availability, 
the \trole{FlightServer} will \textit{reserve a seat}
responds with the price,
to which the \trole{Traveller} confirms or rejects the purchase;
the latter choice will notify the \trole{FlightServer}
to \textit{release the seat}.
Otherwise, the \trole{FlightServer} notifies the
\trole{Traveller} that all flights are full.
The \trole{Traveller} can proceed to query for another
destination.

@@

The current state-of-the-art succeeds in statically
providing communication safety guarantees,
but requires the developer to compromise by adapting
to programming paradigms and practices that are not
conventional for modern web programming.
Developers that use \cite{MVU2020} are required
to implement their web applications in a \textit{functional} 
paradigm using the \textit{Links} 
web programming language \cite{LINKS}.
\cite{PureScript2019} also endorses
functional web programming by requiring developers to
write PureScript \cite{PureScript} applications, but also
enforces user interfaces to be written using 
sequential UI frameworks.

PureScript and Links are not interoperable
with the ever-growing universe of JavaScript libraries for
web development.
It is also unclear whether these tools support idiomatic
practices in event-driven programming, such as callbacks
and asynchronous implementations.
Whilst communication safety guarantees provided by
\cite{PureScript2019,MVU2020} are based on
canonical MPST theory, they are not contextualised
for the dynamic web-based environment.
Going back to the flight booking service as an example,
\cite{PureScript2019,MVU2020} do not handle premature
disconnections; if the \trole{Traveller} receives a quote
and closes their browser prematurely and the \trole{FlightServer}
is not notified of this event, 
the reserved seat will not be released.
These compromises limit the usability of the MPST framework
in modern web programming.

@@

The current state-of-the-art leverage the WebSocket protocol
as the communication transport between roles, which is
simple to reason about and straightforward to relate
to session type theory.
However, WebSocket transport enforces a \emph{server-centric}
network topology, so \cite{PureScript2019,MVU2020}
enforces the same constraint on the range of communication
protocols supported:
server-centric protocols feature exactly one server
role that is involved in all interactions.

This limits the types set of communication topologies
that can be implemented by session-typed web development
proposals. 
Suppose we extend the flight booking service as follow:
we introduce a \trole{TravelAgent} role, which
runs on the browser; the \trole{Traveller} first asks
for a quote from the \trole{TravelAgent} before 
carrying out the usual interactions with the \trole{FlightServer},
so the \trole{Traveller} can compare between both options.
The current state-of-the-art cannot implement this 
extended flight booking service, as the described
communication interactions is no longer server-centric.

>>>>

Consider a \textit{flight booking service}
between a \trole{Traveller} and a \trole{FlightServer}.
The \trole{Traveller} queries the \trole{FlightServer}
about prices to a particular destination.
If there is still availability, 
the \trole{FlightServer} will \textit{reserve a seat}
and respond with the price,
to which the \trole{Traveller} confirms or rejects the purchase;
the latter choice will notify the \trole{FlightServer}
to \textit{release the seat}.
Otherwise, the \trole{FlightServer} notifies the
\trole{Traveller} that all flights are full.
The \trole{Traveller} can proceed to query for another
destination.

@@

The current state-of-the-art succeeds in statically
providing communication safety guarantees,
but requires the developer to compromise by adapting
to programming paradigms and practices that are not
conventional for modern web programming.
Fowler \cite{MVU2020} requires developers
to implement their web applications in a \textit{functional} 
paradigm using the \textit{Links} 
web programming language \cite{LINKS}.
King et al. \cite{PureScript2019} also endorses
functional web programming by requiring developers to
write PureScript \cite{PureScript} applications, but also
enforces user interfaces to be written using 
sequential UI frameworks.

PureScript and Links are not interoperable
with the ever-growing universe of JavaScript libraries for
web development.
It is also unclear whether these tools support idiomatic
practices in event-driven programming, such as callbacks
and asynchronous implementations.
Whilst communication safety guarantees provided by
Fowler \cite{MVU2020} and King et al. \cite{PureScript2019}
are based on
canonical MPST theory, they are not contextualised
for the dynamic web-based environment.
Going back to the flight booking service as an example,
\cite{PureScript2019,MVU2020} do not handle premature
disconnections; if the \trole{Traveller} receives a quote
and closes their browser prematurely and the \trole{FlightServer}
is not notified of this event, 
the reserved seat will not be released.
These compromises limit the usability of the MPST framework
in modern web programming.

@@

The current state of the art leverage the WebSocket protocol
as the communication transport between roles, which is
simple to reason about and straightforward to relate
to session type theory.
However, WebSocket transport enforces a \emph{server-centric}
network topology, so \cite{PureScript2019,MVU2020}
enforces the same constraint on the range of communication
protocols supported:
server-centric protocols feature exactly one server
role that is involved in all interactions.

This limits the types set of communication topologies
that can be implemented by session-typed web development
proposals. 
Suppose we extend the flight booking service as follow:
we introduce a \trole{TravelAgent} role, which
runs on the browser; the \trole{Traveller} first asks
for a quote from the \trole{TravelAgent} before 
carrying out the usual interactions with the \trole{FlightServer},
so the \trole{Traveller} can compare between both options.
The current state of the art cannot implement this 
extended flight booking service, as the described
communication interactions are no longer server-centric.

### 2020-06-18 report/sections/introduction/objectives.tex
Report: final proofread

We tackle the limitations of the current
state-of-the-art with two solutions:

@@

\subparagraph{S2: \newtheory}
We present an extension of the canonical MPST
theory that supports \textit{routed communication}.
We prove that \newtheory also extends the communication safety
properties of the canonical MPST theory, so
processes that implement \newtheory also enjoy
guarantees of \textbf{deadlock-freedom}, \textbf{channel linearity},
and the absence
of \textbf{communication mismatches}.

Through \newtheory, we can express interactions to be
routed via an intermediate endpoint.
By defining an encoding from canonical MPST theory
to \newtheory, we show that we do not lose expressiveness
using routed communication.
We also extend \codegen to implement \newtheory.
\\

By solution \textbf{S1}, we address the limitation \textbf{L1}.
Our work provides the same communication safety guarantees
as the current state-of-the-art, but 
is compatible with modern web programming tools and idioms.
Among the languages that compile to JavaScript,
we select TypeScript as it is arguably the most intuitive
to use, given it is a \textit{superset} of JavaScript 
\cite{UnderstandingTypeScript}.
It provides developers
with type-safety through its gradual, structural type system,
and is also interoperable with existing JavaScript libraries
that the developer may incorporate into their application.
Whilst some \cite{MVU2020} point out this limits its usability for
encoding multiparty session types, we believe that the language
offers sufficient features that we can use to provide developers with
communication safety guarantees whilst preserving a flexible, natural
and idiomatic workflow.
The TypeScript APIs we generate also support idioms commonly
found in modern web development (such as callbacks and
asynchronous operations), and provides an error handling
mechanism tailored to the dynamic web-based environment
(where roles can disconnect prematurely).

>>>>

We tackle the limitations of the current
state of the art with two solutions:

@@

\subparagraph{S2: \newtheory}
We present an extension of the canonical MPST
theory that supports \textit{routed communication}.
We prove that \newtheory also extends the communication safety
properties of the canonical MPST theory, so
processes that implement \newtheory also enjoy
guarantees of \textbf{deadlock-freedom}
and the absence
of \textbf{communication mismatches}.

Through \newtheory, we can express interactions to be
routed via an intermediate endpoint.
By defining an encoding from canonical MPST theory
to \newtheory, we show that we do not lose expressiveness
using routed communication.
We also extend \codegen to implement \newtheory.
\codegen generates APIs that provide \textbf{channel linearity}
by construction, and the extended
implementation to support \newtheory enjoys the same benefits.
\\

By solution \textbf{S1}, we address the limitation \textbf{L1}.
Our work provides the same communication safety guarantees
as the current state of the art, but 
is compatible with modern web programming tools and idioms.
Among the languages that compile to JavaScript,
we select TypeScript as it is arguably the most intuitive
to use, given it is a \textit{superset} of JavaScript 
\cite{UnderstandingTypeScript}.
It provides developers
with type-safety through its gradual, structural type system,
and is also interoperable with existing JavaScript libraries
that the developer may incorporate into their application.
Whilst some \cite{MVU2020} point out that 
this limits its usability for
encoding multiparty session types, we believe that the language
offers sufficient features that we can use to provide developers with
communication safety guarantees whilst preserving a flexible, natural
and idiomatic workflow.
The TypeScript APIs we generate also support idioms commonly
found in modern web development (such as callbacks and
asynchronous operations), and provide an error handling
mechanism tailored to the dynamic web-based environment
(where roles can disconnect prematurely).

### 2020-06-18 report/sections/server/codegen/testing.tex
Report: final proofread

\section{Testing}

The usual unit testing practices from software engineering
apply to our code generator -- the built-in \texttt{unittest} 
package is more than sufficient for our use case. 
Instead, we detail our methodology for \textit{system testing} -- 
executing \fancyname{SessionTS} end-to-end and testing
the generated code.

We implement a test suite on top of \texttt{unittest} APIs
for verifying that \fancyname{SessionTS}
generates \textit{valid TypeScript code}. This is especially useful as
our templates contain both TypeScript syntax and Jinja markup, 
so we cannot easily make sure that each template generates valid code,
let alone checking that the collection of templates generate a valid
TypeScript project altogether.

@@

As the generated TypeScript code makes assumptions about the
environment in which it is used (for example, having the \texttt{ws}
WebSocket package installed on server-side endpoints), we require a
sandbox environment to type-check the generated code. 
The sandbox contains the minimal boilerplate required
for testing -- this involves having the WebSocket package installed
for server-side endpoints, the React.js framework (\cref{chap:react})
instantiated for browser-side endpoints,
and corresponding \texttt{tsconfig.json} files for both targets to
be picked up by the TypeScript Compiler.

>>>>

\section{Testing}
The challenge for testing \codegen
is to verify that the generated code is valid
TypeScript code.
Here, we detail our methodology
for \textit{system testing} -- 
executing \codegen end-to-end and testing
the generated code.

We implement a test suite on top of \texttt{unittest} APIs
for verifying that \codegen
generates \textit{valid TypeScript code}. This is especially useful as
our templates contain both TypeScript syntax and Jinja markup, 
so we cannot easily make sure that each template generates valid code,
let alone checking that the collection of templates generate a valid
TypeScript project altogether.

@@

As the generated TypeScript code makes assumptions about the
environment in which it is used (for example, having the \texttt{ws}
WebSocket package installed on server-side endpoints), we require a
\textit{sandbox environment} to type-check the generated code. 
The sandbox contains the minimal boilerplate required
for testing -- this involves having the WebSocket package installed
for server-side endpoints, the React.js framework (\cref{chap:react})
instantiated for browser-side endpoints,
and corresponding \texttt{tsconfig.json} files for both targets to
be picked up by the TypeScript Compiler.

### 2020-06-18 report/sections/server/extensions/async.tex
Report: final proofread

Callbacks are commonplace in dealing with 
asynchronous APIs in JavaScript
\cite{CallbackHell}; we will go over other widely-used
concurrency primitives shortly.
Whatever the case may be, asynchronous developer implementations
using our generated APIs will not type-check as the function
returns immediately (hence, has return type \lstonelinejs{void}),
even though the callback function has the expected return type.
The developer will encounter the following:

@@

In order to make \fancyname{SessionTS} more relevant and compatible
with modern web programming practices, it is important to
support asynchronous developer implementations as part of our
generated APIs and runtime.

@@

\begin{lstlisting}[language=javascript,numbers=none]
setTimeout(() => console.log('First!', 0);
console.log('Second!');
\end{lstlisting}

@@

const challenger = (key: number) => {
	return new (*@\hl{Promise}@*)((resolve, reject) => {
		keychain.check(key, ok => {
			if (ok)
				resolve(new Implementation.S44('Authorised', ...);
			else 
				resolve(new Implementation.S44('AccessDenied', ...);	
		});
	});
}
\end{lstlisting}
\captionof{lstlisting}{\tprotocol{2FA} \trole{Svr} Implementation using 
Callbacks, \texttt{Promise}s and
\lstonelinejs{async}/\lstonelinejs{await}}
\label{lst:new2fa}
\end{figure}

@@

Generally, minimal changes are required for the runtime.
The key observation is that the concrete types for the any
handlers are the same -- the payload may be wrapped in a Promise
construct.
To handle potential Promises, we wrap existing logic
as a delayed computation, colloquially known as a \textit{thunk}
\cite{Thunk}.
We show an example in \cref{fig:thunk} of how we
adapt the \texttt{performSend()} function generated by
\fancyname{NodeMPST} to support optional \texttt{Promise}s:

@@

				const result = handler(event);
				if (result instanceof Promise) {
					result.then(send);
				} else {
					send(result);
				}
			}		
		}
	}
}
\end{lstlisting}
\captionof{lstlisting}{Generating Runtime Linearity Checks for
React Send State Components}
\label{lst:asynclinearcheck}
\end{figure}

>>>>

Callbacks are commonplace in dealing with 
asynchronous APIs in JavaScript
\cite{CallbackHell}; we will go over other widely-used
concurrency primitives shortly.
Whatever the case may be, asynchronous developer implementations
using our generated APIs will not type-check as the function
returns immediately (hence, has return type \lstonelinejs{void}),
even though the callback function has the expected return type.
The developer will encounter the following type error message:

@@

In order to make \codegen more relevant and compatible
with modern web programming practices, it is important to
support asynchronous developer implementations as part of our
generated APIs and runtime.

@@

\begin{lstlisting}[language=javascript,numbers=none]
setTimeout(() => console.log('First!', 0));
console.log('Second!');
\end{lstlisting}

@@

const challenger = (key: number) => {
	return new (*@\hl{Promise}@*)((resolve, reject) => {
		keychain.check(key, ok => {
			if (ok)
				resolve(new Implementation.S44('Authorised', ...);
			else 
				resolve(new Implementation.S44('AccessDenied', ...);	
		});
	});
}
\end{lstlisting}
\captionof{lstlisting}{Implementing \trole{Svr} in \tprotocol{2FA} using 
Callbacks, \texttt{Promise}s and
\lstonelinejs{async}/\lstonelinejs{await}}
\label{lst:new2fa}
\end{figure}

@@

Generally, minimal changes are required for the runtime.
The key observation is that the ``concrete'' types for the any
handlers are the same; the payload may simply be wrapped in a Promise
construct.
To handle potential Promises, we wrap existing logic
into a delayed computation, colloquially known as a \textit{thunk}
\cite{Thunk}.
We show an example in \cref{fig:thunk} of how we
adapt the \texttt{performSend()} function generated by
\fancyname{NodeMPST} to support optional \texttt{Promise}s:

@@

				const result = handler(event);
				if (result instanceof Promise) {
					result.then(send);
				} else {
					send(result);
				}
			}		
		}
	}
}
\end{lstlisting}
\captionof{lstlisting}{Generating Runtime Linearity Checks for
React Send States}
\label{lst:asynclinearcheck}
\end{figure}
